---
title: "In-Class Template"
author: "Rebecca Martinez"
format: pdf
---

# 1. Setup

## Packages

```{r packages}

# general use and cleaning:

library(tidyverse)

library(here)

library(janitor)

library(snakecase)

# wrap axis labels

library(scales)

# read in .xlsx  files

library(readxl)

# visualize missing data

library(naniar)

# arrange plots

library(patchwork)


```


## Data

```{r data}


# taxonomic info
taxon_list <- read_csv(here("data", "taxon_list.csv")) 

# aquatic inverts
aq_ins <- read_xlsx(here("data", "Aquatic Sampling Data-2026-03-10.xlsx"),
                    sheet = "Aquatic Insects") 

# site info
sites <- read_xlsx(here("data", "Aquatic Sampling Data-2026-03-10.xlsx"),
                   sheet = "sites")

# water quality info
water_quality <- read_xlsx(here("data", "Aquatic Sampling Data-2026-03-10.xlsx"),
                           sheet = "Water Quality",
                           na = c("", "n/a", "N/A", "over", "173+")) 

# time is in decimals when read in with here
# default is to read in many columns as characters
# 
# must manually say what na values are (see above)
# 
readxlsx_ex_water_quality <- read_xlsx(here("data", "Aquatic Sampling Data-2026-03-10.xlsx"),
                              sheet = "Water Quality")



```


# 2. Cleaning

## `sites` object

Goal: create a new object from `sites` that only includes NCOS sites

Use: filter downstream data frames using new object

```{r ncos site object}

# create new object from sites
ncos_sites <- sites |> 
  # filter only include NCOS sites sampled quarterly (4x/year)
  filter(sector == "NCOS" & sampling_frequency == "Quarterly")


```


## Cleaning `taxon_list`

Goal: convert character strings in existing `taxon_list` data frame into snake case

Use: join with aquatic inverts to attach taxonomic information to species IDs

*Note: Janitor turns column names into snake case; (no spaces or capital letters, only underscores)*

```{r taxon list cleaning}

# create new clean object from taxon_list
taxon_list_clean <- taxon_list |> 
  # convert taxon name to snake case
  mutate(
    verbatim_name = to_any_case(verbatim_name,
                                case = "snake"))


```


## Cleaning `water_quality`

Goal: create data frame with median pH, median DO (in mg/L), and median salinity (in ppt) for each date and site

Use: join with aquatic inverts to attach water quality information to surveys

Pseudocode:

- create new object from `water_quality`
- clean column names
- filter join using `semi_join` to only include sites in NCOS sites
- create a new column called `date_site` to join with later data frames using `unite`
- group by `date_site`
- calculate median pH, salinity (ppt), and dissolved oxygen DO (mg/L)
- ungroup data frame
- filter out salinity outliers


```{r water quality cleaning}

# create new clean object from water quality
water_quality_clean <- water_quality |> 
  # clean column names
  clean_names() |> 
  # filtering join use semi: only include sites that appear in ncos_sites (=sites)
  semi_join(ncos_sites, by = "site") |> 
  # create new column to join with later data frames and remove columns after creating new columns
  unite("date_site", date, site, remove = TRUE) |> 
  # group by date_site column
  group_by(date_site) |> 
  # calculate median pH, salinity, DO
  summarize(med_ph = median(p_h, na.rm = TRUE),
            med_sal = median(salinity_ppt, na.rm = TRUE),
            med_do = median(dissolved_oxygen_mg_l, na.rm = TRUE)) |> 
  # ungroup data frame
  ungroup() |> 
  # filter out salinity outliers (salinity = 174.100)
  filter(date_site != "2022-08-12_NVBR") 


```

