---
title: "In-Class Template"
author: "Rebecca Martinez"
format: pdf
---

# 1. Setup

## Packages

```{r packages}

# general use and cleaning:

library(tidyverse)

library(here)

library(janitor)

library(snakecase)

# wrap axis labels

library(scales)

# read in .xlsx  files

library(readxl)

# visualize missing data

library(naniar)

# arrange plots

library(patchwork)


```


## Data

```{r data}


# taxonomic info
taxon_list <- read_csv(here("data", "taxon_list.csv")) 

# aquatic inverts
aq_ins <- read_xlsx(here("data", "Aquatic Sampling Data-2026-03-10.xlsx"),
                    sheet = "Aquatic Insects") 

# site info
sites <- read_xlsx(here("data", "Aquatic Sampling Data-2026-03-10.xlsx"),
                   sheet = "sites")

# water quality info
water_quality <- read_xlsx(here("data", "Aquatic Sampling Data-2026-03-10.xlsx"),
                           sheet = "Water Quality",
                           na = c("", "n/a", "N/A", "over", "173+")) 

# time is in decimals when read in with here
# default is to read in many columns as characters
# 
# must manually say what na values are (see above)
# 
readxlsx_ex_water_quality <- read_xlsx(here("data", "Aquatic Sampling Data-2026-03-10.xlsx"),
                              sheet = "Water Quality")



```


# 2. Cleaning

## `sites` object

Goal: create a new object from `sites` that only includes NCOS sites

Use: filter downstream data frames using new object

```{r ncos site object}

# create new object from sites
ncos_sites <- sites |> 
  # filter only include NCOS sites sampled quarterly (4x/year)
  filter(sector == "NCOS" & sampling_frequency == "Quarterly")


```


## Cleaning `taxon_list`

Goal: convert character strings in existing `taxon_list` data frame into snake case

Use: join with aquatic inverts to attach taxonomic information to species IDs

*Note: Janitor turns column names into snake case; (no spaces or capital letters, only underscores)*

```{r taxon list cleaning}

# create new clean object from taxon_list
taxon_list_clean <- taxon_list |> 
  # convert taxon name to snake case
  mutate(
    verbatim_name = to_any_case(verbatim_name,
                                case = "snake"))


```


## Cleaning `water_quality`

Goal: create data frame with median pH, median DO (in mg/L), and median salinity (in ppt) for each date and site

Use: join with aquatic inverts to attach water quality information to surveys

Pseudocode:

- create new object from `water_quality`
- clean column names
- filter join using `semi_join` to only include sites in NCOS sites
- create a new column called `date_site` to join with later data frames using `unite`
- group by `date_site`
- calculate median pH, salinity (ppt), and dissolved oxygen DO (mg/L)
- ungroup data frame
- filter out salinity outliers


```{r water quality cleaning}

# create new clean object from water quality
water_quality_clean <- water_quality |> 
  # clean column names
  clean_names() |> 
  # filtering join use semi: only include sites that appear in ncos_sites (=sites)
  semi_join(ncos_sites, by = "site") |> 
  # create new column to join with later data frames and remove columns after creating new columns
  unite("date_site", date, site, remove = TRUE) |> 
  # group by date_site column
  group_by(date_site) |> 
  # calculate median pH, salinity, DO
  summarize(med_ph = median(p_h, na.rm = TRUE),
            med_sal = median(salinity_ppt, na.rm = TRUE),
            med_do = median(dissolved_oxygen_mg_l, na.rm = TRUE)) |> 
  # ungroup data frame
  ungroup() |> 
  # filter out salinity outliers (salinity = 174.100)
  filter(date_site != "2022-08-12_NVBR") 


```

## Cleaning `aq_ins`

Goal: create long format data frame in which each row contains the counted individuals in each taxon during each survey on each date with all the taxonomic information for the taxon and the water quality information for the survey

Use: visualize relationships between water quality parameters and aquatic invertebrate abundances

Pseudocode:

- create new object from `aq_ins`
- clean column names
- filter to only include sites in NCOS sites
- select columns of interest
- filter to only include "filtered beaker" samples
- convert the data frame to long format
- group by date, site, taxon
- calculate average abundance per liter (sum abundance divided by 7.5 liters)
- create a new column called `date_site` to join with water quality
- join with water quality
- join with taxon list


*Note: turnong data to long format often makes impplicit (missing) data explicit*

```{r aq ins cleaning}

# create new clean object from aquatic inverts
aq_ins_clean <- aq_ins |> 
  # clean column names
  clean_names() |> 
  # filtering join: only include sites occurring in ncos_sites
  semi_join(ncos_sites, by = c("site")) |> 
  # rename columns
  rename(date = date_on_vial) |> 
  # select columns of interest
  select(date, sample_type, site,
         ostracod,
         copepod,
         hemiptera_corixidae_boatman,
         cladocera,
         nematode,
         diptera_ceratopogonidae, 
         annelida_oligochaete,
         diptera_chironomid,
         annelida_polychaete) |> 
  # filter for only "filtered beaker" samples
  filter(sample_type %in% c("FB 250", "FB250")) |> 
  # convert data frame to long format
  pivot_longer( # columns to make longer
    cols = ostracod:annelida_polychaete,
               # new column for old column names
               names_to = "taxon",
               # new column for how many of each of these
               values_to = "abundance") |> 
  # group by date, site, taxon
  group_by(date, site, taxon) |> 
  # calculate average abundance per liter (sum abundance divided by 7.5 liters)
  summarize(ave_lit = sum(abundance, na.rm = TRUE)/7.5) |> 
  # ungroup data frame
  ungroup() |> 
  # create a new column called `date_site` to join with water quality
  unite("date_site", date, site, remove = FALSE) |> 
  # join with water quality
  left_join(water_quality_clean, by = c("date_site")) |> 
  # join with taxon list
  left_join(taxon_list_clean, by = c("taxon" = "verbatim_name")) |> 
  # add full names of sites
  mutate(site_full = case_when(
    site == "NVBR" ~ "North of Venoco Bridge",
    site == "NEC" ~ "South of Venoco Bridge",
    site == "NMC" ~ "Main Channel",
    site == "NPB" ~ "South of Phelps Bridge",
    site == "NPB1" ~ "North of Phelps Bridge",
    site == "NPB2" ~ "Phelps Road",
    site == "NWP" ~  "West Pond",
    site == "NDC" ~ "Devereux Creek"
  )) |> 
  # set factor levels for site
  mutate(site_full = fct_reorder(site_full, med_sal, .fun = "median", .na_rm = TRUE),
         site_full = fct_rev(site_full))


```

**stopped at unite (above) in class**
# 3. Visualizing

## Visualizing missing values

```{r missingness}

# check for missing data

gg_miss_var(aq_ins_clean)

```

**Which missing values likely matter?**

## Visualizing salinity

Create a visualization depicting site differences in salinity.

- x-axis: `site`
- y-axis: `mean_sal`
- geometry to show salinity during survey: `geom_jitter()`
- geometry to show mean salinity: `stat_summary()`

```{r salinity visual}

# base layer
salinity <- ggplot(data = aq_ins_clean,
       # x-axis: site
       # y-axis: mean salinity
       mapping = aes(x = site_full,
                     y = med_sal)) +
  # first layer: salinity measurements at each survey
  geom_jitter(height = 0,
              width = 0.1,
              shape = 21) +
  # wrap x-axis text
  scale_x_discrete(labels = label_wrap(width = 15)) +
  # second layer: points to represent means
  stat_summary(fun = median,
               color = "red",
               size = 1) +
  # relabel x- and y-axes
  labs(x = "Site",
       y = "Median salinity (ppt)",
       title = "Site salinity") +
  # different theme
  theme_bw()

# display object
salinity


```

**Do sites differ in salinity?**

## Visualizing differences in copepod abundance

- x-axis: site
- y-axis: average abundance/L
- geometry to represent observations: `geom_jitter()`
- geometry to represent medians: `stat_summary()`

Initial cleaning:

```{r}
#| label: copepod-cleaning

# creating new object from clean aquatic inverts 
copepods <- aq_ins_clean |> 
  # filter to only include copepods
  filter(taxon == "copepod") |> 
  # log + 1 transform mean abundance per liter
  mutate(log_ave_lit = log(ave_lit + 1)) 
```

Visualizing differences in copepod abundance across sites:

```{r}
#| label: visualizing-copepods
#| fig-width: 10
#| fig-height: 6

# base layer
copepods <- ggplot(data = copepods,
                   # x-axis: site
                   # y-axis: mean abundance/L
                   mapping = aes(x = site_full,
                                 y = log_ave_lit)) +
  # first layer: points to represent observations
  geom_jitter(height = 0,
              shape = 21,
              width = 0.1) +
  # second layer: points to represent medians
  stat_summary(geom = "point",
               fun = "median",
               size = 4,
               color = "blue") +
  # wrapping x-axis text
  scale_x_discrete(labels = label_wrap(15)) +
  # relabelling x- and y-axis, adding title
  labs(x = "Site",
       y = "log + 1 transformed average abundance/L",
       title = "Copepods") +
  # different theme
  theme_bw()

# displaying object
copepods
```

**Does copepod abundance differ across sites?**

```{r}
#| label: site-diff-plots
#| fig-width: 8
#| fig-height: 10

salinity / copepods
```

**Are differences in median copepod abundance across sites _similar_ to differences in median salinity across sites?**






















