---
title: "Individual Assignment 5"
author: "Rebecca Martinez"
date: "11 May 2026"
format:
  pdf:
    fig-width: 10
    fig-height: 8
    code-overflow: wrap
editor: source
execute:
  message: false
  warning: false
---

# 1. Setup

## Packages

```{r packages}

# general data wrangling and plotting
library(tidyverse)

# file paths
library(here)

# clean column names
library(janitor)

# convert taxon names to snake case
library(snakecase)

# wrap long axis labels
library(scales)

# read Excel files
library(readxl)

# combine plots
library(patchwork)

# custom colors
library(viridis)

```

## Data

```{r data}

# read in taxonomic information
taxon_list <- read_csv(here("data", "taxon_list.csv")) 

# read in aquatic invertebrate data
aq_ins <- read_xlsx(here("data", "Aquatic Sampling Data-2026-03-10.xlsx"),
                    sheet = "Aquatic Insects") 

# read in site information
sites <- read_xlsx(here("data", "Aquatic Sampling Data-2026-03-10.xlsx"),
                   sheet = "sites")

# read in water quality data
# na values are included so unusual entries are treated as missing
water_quality <- read_xlsx(here("data", "Aquatic Sampling Data-2026-03-10.xlsx"),
                           sheet = "Water Quality",
                           na = c("", "n/a", "N/A", "over", "173+")) 

```

# 2. Cleaning

## NCOS sites

```{r ncos sites}

# create object with only NCOS quarterly sampling sites
ncos_sites <- sites |> 
  
  # keep sites in the NCOS sector that were sampled quarterly
  filter(sector == "NCOS" &
           sampling_frequency == "Quarterly")

```

## Taxon list

```{r taxon list cleaning}

# clean taxon names so they match the aquatic invertebrate data
taxon_list_clean <- taxon_list |> 
  
  # convert verbatim taxon names to snake case
  mutate(verbatim_name = to_any_case(verbatim_name,
                                     case = "snake"))

```


## Water quality

```{r water quality cleaning}

# create cleaned water quality data frame
water_quality_clean <- water_quality |> 
  
  # clean column names
  clean_names() |> 
  
  # keep only NCOS quarterly sampling sites
  semi_join(ncos_sites, by = "site") |> 
  
  # create a shared date-site column for joining later
  unite("date_site",
        date,
        site,
        remove = TRUE) |> 
  
  # group by each date and site combination
  group_by(date_site) |> 
  
  # calculate median water quality values for each date-site survey
  summarize(med_ph = median(p_h, na.rm = TRUE),
            med_sal = median(salinity_ppt, na.rm = TRUE),
            med_do = median(dissolved_oxygen_mg_l, na.rm = TRUE)) |> 
  
  # ungroup data frame
  ungroup() |> 
  
  # remove known salinity outlier
  filter(date_site != "2022-08-12_NVBR")

```

## Aquatic invertebrates

```{r aq ins cleaning}

# create cleaned aquatic invertebrate data frame
aq_ins_clean <- aq_ins |> 
  
  # clean column names
  clean_names() |> 
  
  # keep only NCOS quarterly sampling sites
  semi_join(ncos_sites, by = "site") |> 
  
  # rename date column for easier use
  rename(date = date_on_vial) |> 
  
  # keep date, sample type, site, and major invertebrate taxon columns
  select(date,
         sample_type,
         site,
         ostracod,
         copepod,
         hemiptera_corixidae_boatman,
         cladocera,
         nematode,
         diptera_ceratopogonidae, 
         annelida_oligochaete,
         diptera_chironomid,
         annelida_polychaete) |> 
  
  # keep only filtered beaker samples
  filter(sample_type %in% c("FB 250", "FB250")) |> 
  
  # convert taxon columns into long format
  pivot_longer(cols = ostracod:annelida_polychaete,
               names_to = "taxon",
               values_to = "abundance") |> 
  
  # group by date, site, and taxon
  group_by(date, site, taxon) |> 
  
  # calculate average abundance per liter
  summarize(ave_lit = sum(abundance, na.rm = TRUE) / 7.5,
            .groups = "drop") |> 
  
  # create shared date-site column for joining with water quality
  unite("date_site",
        date,
        site,
        remove = FALSE) |> 
  
  # join water quality data
  left_join(water_quality_clean, by = "date_site") |> 
  
  # join taxonomic information
  left_join(taxon_list_clean, by = c("taxon" = "verbatim_name")) |> 
  
  # create full site names
  mutate(site_full = case_when(
    site == "NVBR" ~ "North of Venoco Bridge",
    site == "NEC" ~ "South of Venoco Bridge",
    site == "NMC" ~ "Main Channel",
    site == "NPB" ~ "South of Phelps Bridge",
    site == "NPB1" ~ "North of Phelps Bridge",
    site == "NPB2" ~ "Phelps Road",
    site == "NWP" ~ "West Pond",
    site == "NDC" ~ "Devereux Creek"
  )) |> 
  
  # reorder sites by median salinity
  mutate(site_full = fct_reorder(site_full,
                                 med_sal,
                                 .fun = "median",
                                 .na_rm = TRUE),
         site_full = fct_rev(site_full))

```

# 3. Class figures needed for assignment

## Site salinity

```{r salinity visual}

# create object from aq_ins_clean
salinity <- aq_ins_clean |>
  
  # remove rows only if median salinity is missing
  filter(!is.na(med_sal)) |>
  
  # create plot
  ggplot(mapping = aes(x = site_full,
                       y = med_sal)) +
  
  # show individual salinity measurements for each survey
  geom_jitter(height = 0,
              width = 0.1,
              shape = 21) +
  
  # wrap long site names on x-axis
  scale_x_discrete(labels = label_wrap(width = 15)) +
  
  # show median salinity for each site
  stat_summary(fun = median,
               geom = "point",
               color = "red",
               shape = 18,
               size = 4) +
  
  # label axes and add title
  labs(x = "Site",
       y = "Median salinity (ppt)",
       title = "Site salinity") +
  
  # use theme different from default
  theme_bw() +
  
  # angle x-axis text for readability
  theme(axis.text.x = element_text(angle = 45,
                                   hjust = 1),
        axis.title.x = element_blank())

```

## Copepod abundance

```{r copepod cleaning}

# create data frame for copepod abundance
copepod_data <- aq_ins_clean |> 
  
  # keep only copepods
  filter(taxon == "copepod") |> 
  
  # log + 1 transform average abundance per liter
  mutate(log_ave_lit = log(ave_lit + 1))

```

```{r copepod visual}

# create copepod plot object
copepods <- ggplot(data = copepod_data,
                   
                   # x-axis: site
                   # y-axis: log-transformed average abundance per liter
                   mapping = aes(x = site_full,
                                 y = log_ave_lit)) +
  
  # show individual average abundance/L measurements
  geom_jitter(height = 0,
              width = 0.1,
              shape = 21) +
  
  # show median abundance for each site
  stat_summary(fun = median,
               geom = "point",
               color = "blue",
               shape = 18,
               size = 4) +
  
  # wrap long site names on x-axis
  scale_x_discrete(labels = label_wrap(width = 15)) +
  
  # label axes and add title
  labs(x = "Site",
       y = "log + 1 transformed average abundance/L",
       title = "Copepods") +
  
  # use theme different from default
  theme_bw() +
  
  # angle x-axis text for readability
  theme(axis.text.x = element_text(angle = 45,
                                   hjust = 1),
        axis.title.x = element_blank())


```
  

# 4. Problems

## 1. Visualizing differences across sites in major taxon abundance

### a. Plan the figure

Goal: create a plot showing differences between sites in ostracod abundance.

- x-axis: `site_full`
- y-axis: `log_ave_lit` (log + 1 transformed average abundance/L)
- geometry to represent average abundance/L: jittered points using `geom_jitter()`
- geometry to represent medians: larger colored diamond points using `stat_summary(fun = median)`


### b. Wrangle data

```{r ostracod wrangling}

# create data frame for ostracod abundance
ostracod_data <- aq_ins_clean |>
  
  # keep only ostracods
  filter(taxon == "ostracod") |>
  
  # log + 1 transform average abundance per liter
  mutate(log_ave_lit = log(ave_lit + 1))

# display 10 random rows from ostracod data frame
ostracod_data |>
  slice_sample(n = 10)

```


### c. Code figure

```{r ostracod visual}

# create ostracod plot object
ostracods <- ggplot(data = ostracod_data,
                    
                    # x-axis: site
                    # y-axis: log-transformed average abundance per liter
                    mapping = aes(x = site_full,
                                  y = log_ave_lit)) +
  
  # show individual average abundance/L measurements
  geom_jitter(height = 0,
              width = 0.1,
              shape = 21) +
  
  # show median abundance for each site with a visually distinct point
  stat_summary(fun = median,
               geom = "point",
               color = "purple",
               shape = 18,
               size = 4) +
  
  # wrap long site labels
  scale_x_discrete(labels = label_wrap(width = 15)) +
  
  # label axes and add title
  labs(x = "Site",
       y = "log + 1 transformed average abundance/L",
       title = "Ostracods") +
  
  # use theme different from default
  theme_bw() +
  
  # angle x-axis text for readability
  theme(axis.text.x = element_text(angle = 45,
                                   hjust = 1),
        axis.title.x = element_blank())

```


### d. Create multipanel figure

```{r multipanel figure}

# combine salinity, copepod, and ostracod plots
salinity / (copepods | ostracods)

```


### e. Interpretation

**Differences between sites in copepod and ostracod abundance:**
Copepod abundance is relatively high at South of Venoco Bridge, South of Phelps Bridge, and Devereux Creek, shown by the higher blue median points. Ostracod abundance is highest at South of Phelps Bridge and Devereux Creek, where the purple median points are higher than the other sites and several individual observations are also high.

**Relationship between ostracod abundance and salinity:**
Ostracod abundance appears to be higher at lower-salinity sites rather than higher-salinity sites. This is shown by the low red median salinity points at South of Phelps Bridge and Devereux Creek, where ostracod medians are highest, while the Venoco Bridge sites have higher salinity but lower ostracod medians.

  
## 2. Visualizing total abundance as a function of salinity

### a. Plan the figure

Goal: create a plot showing log-transformed average abundance/L as a function of salinity for each taxon.

- x-axis: `med_sal` (median salinity in ppt)
- y-axis: `log_ave_lit` (log + 1 transformed average abundance/L)
- color: `taxon_title`
- geometry: points using `geom_point()`
- facets: `facet_wrap(~ taxon_title, scales = "free")`


### b. Wrangle data

```{r abundance salinity wrangling}

# create object for abundance and salinity plot
abundance_salinity <- aq_ins_clean |>
  
  # remove rows only if median salinity is missing
  filter(!is.na(med_sal)) |>
  
  # create cleaned columns needed for plotting
  mutate(
    
    # log + 1 transform average abundance per liter
    log_ave_lit = log(ave_lit + 1),
    
    # make taxon names easier to read
    taxon_title = str_replace_all(taxon, "_", " "),
    
    # convert taxon names to title case
    taxon_title = str_to_title(taxon_title)
  ) |>
  
  # order taxa by maximum abundance
  mutate(taxon_title = fct_reorder(taxon_title,
                                   ave_lit,
                                   .fun = max,
                                   .desc = TRUE)) |>
  
  # keep only columns needed for figure
  select(date,
         site,
         site_full,
         med_sal,
         taxon,
         taxon_title,
         ave_lit,
         log_ave_lit)

# display 10 random rows from data frame
abundance_salinity |>
  slice_sample(n = 10)

```


### c. Code your figure

```{r abundance salinity figure}

# create plot showing abundance as a function of salinity
abundance_salinity_plot <- ggplot(data = abundance_salinity,
                                  
                                  # x-axis: median salinity
                                  # y-axis: log-transformed average abundance per liter
                                  # color by taxon
                                  mapping = aes(x = med_sal,
                                                y = log_ave_lit,
                                                color = taxon_title)) +
  
  # show individual abundance/salinity observations
  geom_point() +
  
  # create separate panels for each taxon with free x- and y-axis scales
  facet_wrap(~ taxon_title,
             scales = "free") +
  
  # use non-default color palette
  scale_color_viridis_d(option = "turbo") +
  
  # label x- and y-axes
  labs(x = "Median salinity (ppt)",
       y = "log + 1 transformed average abundance/L",
       title = "Invertebrate Abundance as a Function of Salinity") +
  
  # use theme different from default
  theme_bw() +
  
  # remove legend because taxa are labeled in facet titles
  theme(legend.position = "none")

# display plot
abundance_salinity_plot

```


### d. Interpretation

Most taxa have higher abundance at low salinity, shown by points clustered on the left side of the panels where median salinity is around 0–10 ppt. Copepods and ostracods have the highest log-transformed abundance overall, but copepods occur across a wider range of salinity values. Ostracods appear more abundant at lower salinity and lower at higher salinity, suggesting abundance may decrease as salinity increases. Other taxa, like nematodes and annelids, stay low across most salinity values, with points clustered near zero on the y-axis.



